---
title: "Social Familiarity and Reinforcement Value"
subtitle: "Publication-aligned analysis in R"
author: "Haoran Wan"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
engine: knitr
---

## Purpose and scientific scope

This report reproduces the nonlinear demand analyses reported by Schulingkamp, Wan, and Hackenberg (2023). Four focal rats completed fixed-ratio (FR) schedules for 10-, 30-, and 60-second access to a cagemate or non-cagemate. The primary outcome was the rate of obtained social interactions during active session time.

The workflow follows the published analysis: fit 24 subject-condition zero-bounded exponential (ZBEn) demand curves, summarize model fit and derived demand parameters, compare familiarity with aggregate contrasts, and compare duration with nested nonlinear models. Exploratory models are outside the scope of this publication replication.

> **Data access.** Download the analysis-ready data from the [OSF record found by the exact article title](https://osf.io/search/?q=%22Social%20familiarity%20and%20reinforcement%20value%22) and save it as `Analysis/Table1.csv`. Alternatively, set the `SOCIAL_DEMAND_DATA` environment variable to the downloaded file.

## Setup

```{r setup}
required_packages <- c("dplyr", "knitr", "minpack.lm", "readr", "tidyr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    "Install missing packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(dplyr)
library(minpack.lm)
library(readr)
library(tidyr)

options(digits = 4, scipen = 4)
set.seed(20230122)

ihs <- function(x) {
  log10(0.5 * x + sqrt(0.25 * x^2 + 1))
}

inverse_ihs <- function(y) {
  (10^(2 * y) - 1) / 10^y
}

zben <- function(fr, log_alpha, q0) {
  ihs_q0 <- ihs(q0)
  ihs_q0 * exp((-exp(log_alpha) / ihs_q0) * q0 * fr)
}

essential_value <- function(alpha) {
  1 / (100 * alpha)
}

nls_control <- nls.lm.control(maxiter = 1024, maxfev = 100000)
```

## Data import and validation

The OSF file is kept outside version control. The loader checks an explicit environment variable first and then the two paths expected when rendering from the repository root or the `Analysis` directory.

```{r data-import}
data_candidates <- unique(c(
  Sys.getenv("SOCIAL_DEMAND_DATA", unset = NA_character_),
  "Analysis/Table1.csv",
  "Table1.csv"
))
data_candidates <- data_candidates[!is.na(data_candidates) & nzchar(data_candidates)]
existing_data <- data_candidates[file.exists(data_candidates)]

if (length(existing_data) == 0L) {
  stop(
    "Data not found. Download Table1.csv from OSF and place it in Analysis/, ",
    "or set SOCIAL_DEMAND_DATA to the downloaded file."
  )
}

data_path <- existing_data[[1L]]
expected_columns <- c(
  "Rat", "Social Familiarity", "Social Duration", "Social FR",
  "Interaction Rate"
)

raw_data <- read_csv(data_path, show_col_types = FALSE)
missing_columns <- setdiff(expected_columns, names(raw_data))

if (length(missing_columns) > 0L) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

if (anyNA(raw_data[expected_columns])) {
  stop("Required analysis fields contain missing values.")
}

if (any(raw_data$`Interaction Rate` < 0) || any(raw_data$`Social FR` <= 0)) {
  stop("Interaction rates must be nonnegative and FR prices must be positive.")
}

observed_design <- list(
  rats = sort(unique(raw_data$Rat)),
  familiarity = sort(unique(raw_data$`Social Familiarity`)),
  duration = sort(unique(raw_data$`Social Duration`))
)

if (length(observed_design$rats) != 4L ||
    !identical(observed_design$familiarity, c("Cagemate", "Non-cagemate")) ||
    !identical(observed_design$duration, c("10 Sec", "30 Sec", "60 Sec"))) {
  stop("The data do not contain the expected 4 x 2 x 3 study design.")
}

analysis_data <- raw_data |>
  transmute(
    rat = Rat,
    familiarity = `Social Familiarity`,
    duration = `Social Duration`,
    fr = `Social FR`,
    interaction_rate = `Interaction Rate`,
    lq = ihs(`Interaction Rate`)
  ) |>
  group_by(rat, familiarity, duration, fr) |>
  summarise(
    interaction_rate = mean(interaction_rate),
    lq = mean(lq),
    sessions = n(),
    .groups = "drop"
  )

tibble(
  source = normalizePath(data_path),
  session_rows = nrow(raw_data),
  aggregated_rows = nrow(analysis_data),
  rats = n_distinct(analysis_data$rat),
  curves = n_distinct(interaction(analysis_data$rat, analysis_data$familiarity, analysis_data$duration))
) |>
  knitr::kable(caption = "Validated analysis data")
```

## Subject-condition demand curves

The ZBEn model is fitted independently to each rat, familiarity condition, and duration. `Q0` is predicted consumption at zero price; `alpha` is elasticity; essential value is `1 / (100 alpha)`. `Pmax` and `Omax` are approximated on a 0.1-FR grid.

```{r subject-curves}
fit_curve <- function(curve_data) {
  fit <- nlsLM(
    lq ~ zben(fr, log_alpha, q0),
    data = curve_data,
    start = list(log_alpha = -6, q0 = 50),
    lower = c(log_alpha = -Inf, q0 = .Machine$double.eps),
    control = nls_control
  )

  residuals <- residuals(fit)
  sse <- sum(residuals^2)
  sst <- sum((curve_data$lq - mean(curve_data$lq))^2)
  r_squared <- 1 - sse / sst

  estimates <- coef(fit)
  alpha <- exp(unname(estimates[["log_alpha"]]))
  q0 <- unname(estimates[["q0"]])

  price_grid <- seq(min(analysis_data$fr), max(analysis_data$fr), by = 0.1)
  predicted_lq <- zben(price_grid, log(alpha), q0)
  response_output <- inverse_ihs(predicted_lq) * price_grid
  maximum_index <- which.max(response_output)

  tibble(
    alpha = alpha,
    q0 = q0,
    essential_value = essential_value(alpha),
    pmax = price_grid[[maximum_index]],
    omax = response_output[[maximum_index]],
    r_squared = r_squared,
    converged = isTRUE(fit$convInfo$isConv)
  )
}

curve_keys <- analysis_data |>
  distinct(rat, familiarity, duration) |>
  arrange(rat, familiarity, duration)

curve_results <- lapply(seq_len(nrow(curve_keys)), function(i) {
  key <- curve_keys[i, ]
  curve_data <- analysis_data |>
    filter(
      rat == key$rat,
      familiarity == key$familiarity,
      duration == key$duration
    )

  bind_cols(key, fit_curve(curve_data))
}) |>
  bind_rows()

if (!all(curve_results$converged)) {
  stop("At least one subject-condition demand curve did not converge.")
}

curve_results |>
  mutate(across(c(alpha, q0, essential_value, pmax, omax, r_squared), ~ round(.x, 3))) |>
  knitr::kable(caption = "Subject-condition ZBEn parameter estimates")

curve_fit_summary <- curve_results |>
  summarise(
    curves = n(),
    mean_r_squared = mean(r_squared),
    minimum_r_squared = min(r_squared),
    maximum_r_squared = max(r_squared)
  )

curve_fit_summary |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = "Demand-curve fit summary")
```

## Aggregate comparisons

### Social familiarity

The familiarity analysis collapses across duration and fits a four-parameter model. Separate `alpha` and `Q0` parameters are estimated for the cagemate and non-cagemate conditions. The same condition-specific `Q0` term is used throughout the equation.

```{r familiarity-model}
familiarity_data <- analysis_data |>
  group_by(rat, familiarity, fr) |>
  summarise(lq = mean(lq), .groups = "drop") |>
  mutate(
    is_cagemate = as.numeric(familiarity == "Cagemate"),
    is_non_cagemate = as.numeric(familiarity == "Non-cagemate")
  )

familiarity_fit <- nlsLM(
  lq ~
    ihs(q_cagemate * is_cagemate + q_non_cagemate * is_non_cagemate) *
    exp(
      (-exp(log_alpha_cagemate * is_cagemate + log_alpha_non_cagemate * is_non_cagemate) /
        ihs(q_cagemate * is_cagemate + q_non_cagemate * is_non_cagemate)) *
      (q_cagemate * is_cagemate + q_non_cagemate * is_non_cagemate) * fr
    ),
  data = familiarity_data,
  start = list(
    log_alpha_cagemate = -6,
    log_alpha_non_cagemate = -6,
    q_cagemate = 50,
    q_non_cagemate = 50
  ),
  control = nls_control
)

normal_contrast <- function(fit, weights, label) {
  estimates <- coef(fit)
  weights <- weights[names(estimates)]
  estimate <- sum(weights * estimates)
  standard_error <- sqrt(drop(weights %*% vcov(fit) %*% weights))
  z_value <- estimate / standard_error

  tibble(
    comparison = label,
    estimate = estimate,
    standard_error = standard_error,
    z_value = z_value,
    p_value = 2 * pnorm(-abs(z_value))
  )
}

alpha_weights <- c(
  log_alpha_cagemate = 1,
  log_alpha_non_cagemate = -1,
  q_cagemate = 0,
  q_non_cagemate = 0
)
q0_weights <- c(
  log_alpha_cagemate = 0,
  log_alpha_non_cagemate = 0,
  q_cagemate = 1,
  q_non_cagemate = -1
)

familiarity_results <- bind_rows(
  normal_contrast(familiarity_fit, alpha_weights, "Elasticity (log alpha)"),
  normal_contrast(familiarity_fit, q0_weights, "Demand intensity (Q0)")
)

familiarity_results |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(caption = "Familiarity tests")
```

### Interaction duration

Duration is evaluated with nested nonlinear models. The full model estimates separate `alpha` and `Q0` values for 10, 30, and 60 seconds. Each reduced model constrains one parameter to be equal across durations.

```{r duration-tests}
duration_data <- analysis_data |>
  group_by(rat, duration, fr) |>
  summarise(lq = mean(lq), .groups = "drop") |>
  mutate(
    d10 = as.numeric(duration == "10 Sec"),
    d30 = as.numeric(duration == "30 Sec"),
    d60 = as.numeric(duration == "60 Sec")
  )

duration_full_fit <- nlsLM(
  lq ~ ihs(q10 * d10 + q30 * d30 + q60 * d60) *
    exp(
      (-exp(a10 * d10 + a30 * d30 + a60 * d60) /
        ihs(q10 * d10 + q30 * d30 + q60 * d60)) *
      (q10 * d10 + q30 * d30 + q60 * d60) * fr
    ),
  data = duration_data,
  start = list(a10 = -6, a30 = -6, a60 = -6, q10 = 50, q30 = 50, q60 = 50),
  control = nls_control
)

duration_common_alpha_fit <- nlsLM(
  lq ~ ihs(q10 * d10 + q30 * d30 + q60 * d60) *
    exp(
      (-exp(a) / ihs(q10 * d10 + q30 * d30 + q60 * d60)) *
      (q10 * d10 + q30 * d30 + q60 * d60) * fr
    ),
  data = duration_data,
  start = list(a = -6, q10 = 50, q30 = 50, q60 = 50),
  control = nls_control
)

duration_common_q0_fit <- nlsLM(
  lq ~ ihs(q0) *
    exp(
      (-exp(a10 * d10 + a30 * d30 + a60 * d60) / ihs(q0)) *
      q0 * fr
    ),
  data = duration_data,
  start = list(a10 = -6, a30 = -6, a60 = -6, q0 = 50),
  control = nls_control
)

nested_f_test <- function(full_fit, reduced_fit, label) {
  df_numerator <- df.residual(reduced_fit) - df.residual(full_fit)
  df_denominator <- df.residual(full_fit)
  sse_full <- deviance(full_fit)
  sse_reduced <- deviance(reduced_fit)
  f_value <- ((sse_reduced - sse_full) / df_numerator) /
    (sse_full / df_denominator)

  tibble(
    comparison = label,
    df_numerator = df_numerator,
    df_denominator = df_denominator,
    f_value = f_value,
    p_value = pf(f_value, df_numerator, df_denominator, lower.tail = FALSE)
  )
}

duration_results <- bind_rows(
  nested_f_test(duration_full_fit, duration_common_alpha_fit, "Elasticity (alpha)"),
  nested_f_test(duration_full_fit, duration_common_q0_fit, "Demand intensity (Q0)")
)

duration_results |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  knitr::kable(caption = "Duration tests")
```

## Reproducibility checks

The checks below compare computed results with the expected values from the analysis workflow. A tolerance of .01 accommodates rounding and minor optimizer differences.

```{r validation}
computed_values <- c(
  mean_r_squared = curve_fit_summary$mean_r_squared,
  familiarity_alpha_p = familiarity_results$p_value[[1]],
  familiarity_q0_p = familiarity_results$p_value[[2]],
  duration_alpha_p = duration_results$p_value[[1]],
  duration_q0_p = duration_results$p_value[[2]]
)

expected_values <- c(
  mean_r_squared = 0.91,
  familiarity_alpha_p = 0.0163,
  familiarity_q0_p = 0.471,
  duration_alpha_p = 0.226,
  duration_q0_p = 0.805
)

validation_results <- tibble(
  statistic = names(computed_values),
  computed = unname(computed_values),
  expected = unname(expected_values),
  absolute_difference = abs(computed - expected),
  passes = absolute_difference <= 0.01
)

validation_results |>
  mutate(across(c(computed, expected, absolute_difference), ~ round(.x, 4))) |>
  knitr::kable(caption = "Analysis validation checks")

if (!all(validation_results$passes)) {
  stop("At least one computed result differs materially from the expected value.")
}
```

## Interpretation

Demand for social interaction decreased with FR price, and the ZBEn model described the 24 demand curves well. The familiarity contrast indicates higher log `alpha` for the cagemate condition than for the non-cagemate condition, whereas `Q0` did not differ. Interaction duration did not significantly improve model fit for either `alpha` or `Q0`. These aggregate results should be interpreted cautiously because the study included four focal rats and familiarity was confounded with condition order.

## Session information

```{r session-information}
sessionInfo()
```
